---
title: 'Cyclist: Google DA Capstone'
author: "Jonathan Borrell"
date: "2023-05-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background: Google DA Capstone Project for certification
Elected to take up role as junior data analyst working with marketing analyst team at Cyclistic,
a bike-share company in Chicagp. Cyclistic's program features more than 5,800 bicycles and 600 docking
stations. Cyclistic sets itself apart from its competitors by also offering reclining bikes, hand tricycles and
cargo bikes to those who cannot ride a standard two-wheel bike. The majority of their riders opt for traditional
bikes with about 8% of riders using their alternative assistance options. Cyclistic users are more likely to ride
for leisure, but about 30% use them to commute to work each day. Customers who purchase single-ride or full-day passes
are referred to as casual riders whereas customers who purchase annual memberships are Cyclistic members.

The director of marketing believes the company's future success depends on maximizing the number of annual memberships.
Therefore, my team was tasked in finding out how casual riders and annual members use Cyclistic bikes differently. From
these insights, my team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic
executives must approve these recommendations, so our insights must be polished and backed by compelling data and detail-oriented
clear data vizualizations.

This project will broken up into 6 stages encompassing the Data Analytics stages:

- Ask: Identify the business task and determine the key stakeholders.
- Prepare: Collect the data, identify how it's organized and determine the credibility of the data.
- Process: Select the tool for data cleaning, check for errors and document the cleaning process.
- Analyze: Organize and format the data, aggregate the data so that it's useful, perform calculations and identify trends and
relationships.
- Share: Use design thinking principles and data-driven storytelling approaches, present the findings with effective vizualization.
Ensure the analysis has answered the business task.
- Act: Share teh final conclusion and the recommendations.

### **Ask:**

Business Task:
Recommend marketing strategies aimed at converting casual riders into annual members by better understanding how annual members
and casual riders use Cyclistic bike services differently

Stakeholders:
Lily Moreno: The director of marketing and my current manager.
Cyclistic executive team: A detail-oriented executive team who will decide whether to approve the recommended marketing program
Cyclistic marketing analytics team: A team of data analysts responsible for collecting, analyzing, and reporting data that helps
guide Cyclistic's marketing strategy.


### **Prepare:**

For this project, I will use the public data of Cyclistic's historical trip data to analyze and identify trends. This data is provided
to me through Coursera under the Google Data Analytics Certificate program pulled from a course document link providing this [index](https://divvy-tripdata.s3.amazonaws.com/index.html).

I downloaded the ZIP files containing the .csv files from the link above. I will be uploading this file utilizing R Markdown in R
and filtering the knitted .html file onto Github due to a slight favoritism to the site for version control. I will be analyzing the
previous 12 months worth of data starting from April 2023 and going all of the way back to March 2022.

In the first stages I utilized Google Sheets to try to upload the data but found the amount of information being imported Google Sheets
could not handle. I then switched to Microsoft Excel to at least get a glimpse into the data's structure and find out what data I would be
working with. Information contained in each .csv file was as follows: ride id, rideable type, start and end times, start and end station ids,
latitude and longitude of start and end stations and start and end stations.

### **Process:**

I will be using the R Programming language for most, if not all, of my organization and data cleaning.
I will be using the following R libraries:


```{r}
#Load the tidyverse, lubridate, data.table, readr and ggplot2 packages
##Excellent package for all around cleaning and manipulation
library(tidyverse)

##For customization and manipulation of date/time
library(lubridate)

##For dataframe/table manipulation, also allows for faster read in time
library(data.table)

##For quick read/write in of .csv files
library(readr)

##For plotting/visualization later on
library(ggplot2)
```

I then worked on importing my data and created dataframes to call upon for later rbinding into 1 single data frame.

```{r}
#Set the working directory
setwd("C:/Users/Jon/Documents/Capstone BikeShare Data/tripdata-csv")

##Start with clear environment
rm(list = ls())
```
```{r}
##Assign simple operators df = dataframe 1-12 using data.table::fread()
##Noticed this had fastest output compared to other read options based on tests
##Data frame read in ~15 seconds
df1 <- data.table::fread("202204-divvy-tripdata.csv")
df2 <- data.table::fread("202205-divvy-tripdata.csv")
df3 <- data.table::fread("202206-divvy-tripdata.csv")
df4 <- data.table::fread("202207-divvy-tripdata.csv")
df5 <- data.table::fread("202208-divvy-tripdata.csv")
df6 <- data.table::fread("202209-divvy-tripdata.csv")
df7 <- data.table::fread("202210-divvy-tripdata.csv")
df8 <- data.table::fread("202211-divvy-tripdata.csv")
df9 <- data.table::fread("202212-divvy-tripdata.csv")
df10 <- data.table::fread("202301-divvy-tripdata.csv")
df11 <- data.table::fread("202302-divvy-tripdata.csv")
df12 <- data.table::fread("202303-divvy-tripdata.csv")
```

```{r}
##Verify structure and data types of data frames
str(df1)
str(df2)
str(df3)
str(df4)
str(df5)
str(df6)
str(df7)
str(df8)
str(df9)
str(df10)
str(df11)
str(df12)
```

